A formal approach but practical approach in train regulation

### 2025-06-25 paper/main.tex
Déplacer présentation de la maquette dans l'intro

\begin{wrapfigure}[8]{r}{0.28\textwidth}
 \begin{minipage}{0.28\textwidth}
        \centering
        \vspace{-8mm}
\includegraphics[height=30mm]{maquette.png}
    \end{minipage}
\end{wrapfigure}

\paragraph{Final words on the N-scale model.} Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm.  We operate  three trains in parallel. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train will be able to stop before it reaches the end of the block. In practice, we stop the trains in the middle.

@@

\paragraph{Outline.}
In Section~\ref{sec:model_configurations}, we introduce the main building blocks of our model: network infrastructure, trains and the regulator. In Section~\ref{sec:model_semantics}, we present the dynamic aspects of our system: how trains can move and how the regulator schedules them. We conclude this section by stating what we expect of correct networks. Then, in Section~\ref{sec:tla-formalisation}, we build a TLA+ implementation of our model and its dynamics, which we use to check network correctness. Thus, model checking a schedule \emph{certifies} its correctness. Our last contribution, in Section~\ref{sec:composition}, explains how to build large schedules by composing smaller ones. While we don't prove the correctness \emph{by construction} of composed schedules, we use the TLA+ implementation to certify, \emph{a posteriori}, the correctness of composed schedules.

@@

The main principle behind our scheduling process is based on the notion of route. A route for a train is a sequence $\vec{x} = x_1, \ldots, x_k$ of (contiguous) blocks and switches starting and ending on a block.  A \emph{global routing}, denoted $X = (\vec{x}^{\tid{id}})_{\tid{id} \in \trains}$, is a route for each train.  Two routes $\vec{x}$ and $\vec{y}$ have a critical resource if  $x_i = y_j$ for some $i,j$. We denote by $k_{\tid{id}}$ the index of the last component of sequence $\vec{x}^\tid{id}$ for any $\tid{id}\in \trains$.

Informally, a \emph{safe step} is a global routing  such that there are no critical resources between two routes and there is a network configuration $\sigma$ such that  for each train $\tid{id}$, there is a direction according to which starting at $x_1^{\tid{id}}$, following that direction and given the configuration $\sigma$, one reaches $x_{k_{\tid{id}}}^{\tid{id}}$. Let us turn to the technical definition.

Given a route $\vec{x} = x_1, \ldots, x_k$, we say that $\vec{x}$ has a privileged pair denoted $\text{priv}(\vec{x}) = \langle d, \sigma\rangle \in \directions \times \Sigma$ whenever for any pair $i < j$ such that
\begin{itemize}
\item  $x_i$, $x_j$ are blocks and  
\item $x_{i+1}, \ldots, x_{j-1}$ are switches
\end{itemize} 
then  $\sucblock(x_i, d, \sigma) = x_j$. A safe step is a global routing $X$ such that:
\begin{itemize}
\item for any $\tid{id} \neq \tid{id'} \in \trains$, $\vec{x}^{\tid{id}}$ and $\vec{x}^{\tid{id'}}$ have no critical resources,
\item for any $\tid{id} \in \trains$, there is a privileged pair $\langle d, \sigma\rangle$  for $\vec{x}^{\tid{id}}$,
\item for any $\tid{id} \in \trains$, any component $x_j^\tid{id}$ with $j \leq k_{\tid{id}}$ occurs only once in $\vec{x}^{\tid{id}}$,
\item all the privileged pairs share the same value $\sigma$. 
\end{itemize}

Any global routing can be seen as the composition of safe steps and actually, that provides a scheduling algorithm. We say that a sequence of blocks indexed by trains $\vec{\bid{}}_{\tid{id} \in \trains}$ is \emph{separate} if there are no two trains $\tid{id} \neq \tid{id'}$ with $\vec{\bid{}}[\tid{id}] = \vec{\bid{}}[\tid{id'}]$. This notion induces a graph: let $G = \langle V, E\rangle$ with $V$ the set of separate positions and $E$ the set of safe steps between them.  Then, to compute global routes, we look for a path in the graph from the source to the target. The  $A^*$ algorithm we implemented is proving highly effective. This strategy has a main advantage, it is quite flexible with respect to new events. Suppose that at some point a train is blocked, we may recompute safe steps for the other trains, possibly releasing a deadlock.

@@

To each safe step $X = (\vec{x}_i)_{i \in \trains}$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$. $S(C)$ is next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\sigma$ during the initial step  and we lock signals to red for each train's destination in its intended direction.  
\begin{itemize}
\item for each train $\tid{id}$, $\trainSeq[id] = \trainTuple{\tid{id}}{\bid{id}}{d}{\su{d}{x^{\tid{id}}_{k_{\tid{id}}}}}$ with $\langle d,\sigma\rangle = \text{priv}(\vec{x}^{\tid{id}})$,
\item $R = \langle E,  \bid{id} \in \blocks \mapsto 0, \emptyset\rangle$, $E$ being defined just below,
\item $\sigma$ is the unique network configuration shared by all privileged pairs (here we suppose that there is at least one(!) train),
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$: $\langle \bid{id}, d\rangle \mapsto \siggreen$,
\end{itemize}

Now, let us turn to $E$. Let $\sigma$ be the unique  shared network configuration in privileged pairs and let 
$\vec{y} = \{ \langle x_{k_{\tid{id}}}^\tid{id}, d\rangle \mid \tid{id} \in \trains \text{ and }  \langle d, \sigma\rangle = \text{priv}(\vec{x}^\tid{id}) \}$. 
We define $E(\bot) = (\turn{\sid{id}}{\sigma(\sid{id})})_{\sid{id} \in \turnouts} \cdot (\text{signalred}(x,d))_{(x,d) \in \vec{y}} $.

Given some train $\tid{id} \in \trains$, let $B = (j \leq k_\tid{id} \mid  x_j^\tid{id} \text{ is a block})$. 
Let $\text{priv}(\vec{x}^\tid{id}) = \langle d, \sigma\rangle$, then, $E( \tid{id} ) = \langle d,  (\emptyList)_{j \in B}\rangle$.

@@

Two states $S$ and $S'$ are said to be \emph{compatible} whenever  for any train $\tid{id}$, we have $\mathtt{dst}(S, \tid{id})  =  \mathtt{src}(S', \tid{id})$. In other words, trains must start in $S'$ at destinations of $S$. We define the composition only compatible states. Actually, since we are interested by safe steps compositions, we restrict our attention to their compositions.

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}$. Let $S = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \regulator, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \regulator', \switches', \signals'\rangle \rangle$ as follows:

\begin{enumerate}
\item for all train $\tid{id} \in \trains$, let $P_0 = S.\Gamma[\tid{id}].P$ and $P_k = S_k.\Gamma[\tid{id}].P$, then $\Gamma'[tid] = \Gamma[\tid{id}][ P \leftarrow P_0 \cdot P_k]$;
\item $\sigma' = \sigma$,
\item $F' = F$,
\end{enumerate}

For $1)$, in other words, we concatenates the train orders without changing its position, nor its direction. For $2)$, we start as we would do for $S$. The same for $3)$.

>>>>

@@

\paragraph{N-scale model.}
\begin{wrapfigure}[10]{r}{0.28\textwidth}
	\begin{minipage}{0.28\textwidth}
		\centering
		\vspace{-8mm}
		\includegraphics[height=30mm]{maquette.png}
		\caption{The N-scale model.}
	\end{minipage}
\end{wrapfigure}
We tested the work presented in this paper on a real-world N-scale train model. Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm. We operate three trains concurrently. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train are able to stop before reaching the end of the block.

\paragraph{Outline.}
In Section~\ref{sec:model_configurations}, we introduce the main building blocks of our model: network infrastructure, trains and the regulator. In Section~\ref{sec:model_semantics}, we present the dynamic aspects of our system: how trains can move and how the regulator schedules them. We conclude this section by stating what we expect of correct networks. Then, in Section~\ref{sec:tla-formalisation}, we present a TLA+ implementation of our model, which we use to check network correctness: model checking a schedule \emph{certifies} its correctness. Our last contribution, in Section~\ref{sec:composition}, explains how to build large schedules by composing smaller ones. While we don't prove the correctness \emph{by construction} of composed schedules, we use the TLA+ implementation to certify, \emph{a posteriori}, the correctness of composed schedules.

@@

The main principle behind our scheduling process is based on the notion of route. A route for a train is a sequence $\vec{x} = x_1, \ldots, x_k$ of blocks such that, for each $i < k$, there is a network configuration $\sigma$  and a direction $d$ with $\sucblock(x_i, d, \sigma) = x_{i+1}$.  The values $d$ and $\sigma$ are the witness of the transition. A \emph{global routing}, denoted $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$, is a route for each train.  Two routes $\vec{x}$ and $\vec{y}$ have a critical resource if  $x_i = y_j$ for some $i,j$. We denote by $k_{\tid{i}}$ the index of the last component of sequence $\vec{x}^\tid{i}$ for any $\tid{i}\in \trains$.

A \emph{safe step} is a global routing  coming with a network configuration $\sigma$ and a direction map $d: \trains \to \directions$ such that:
\begin{itemize}
\item there are no critical resources between any two routes,
\item $\sigma$ and $d(\tid{i})$ are witnesses of all transitions in $\vec{x}^\tid{i}$. 
\item for any $\tid{i} \in \trains$, any component $x_j^\tid{i}$ with $j \leq k_{\tid{i}}$ occurs only once in $\vec{x}^{\tid{i}}$,
\end{itemize}

Any global routing can be seen as the composition of safe steps and actually, that provides a scheduling algorithm. We say that a sequence of blocks indexed by trains $\vec{\bid{}}_{\tid{i} \in \trains}$ is \emph{separate} if there are no two trains $\tid{i} \neq \tid{id'}$ with $\vec{\bid{}}[\tid{i}] = \vec{\bid{}}[\tid{id'}]$. This notion induces a graph: let $G = \langle V, E\rangle$ with $V$ the set of separate positions and $E$ the set of safe steps between them.  Then, to compute global routes, we look for a path in the graph from the source to the target. The  $A^*$ algorithm we implemented is proving highly effective. This strategy has a main advantage, it is quite flexible with respect to new events. Suppose that at some point a train is blocked, we may recompute safe steps for the other trains, possibly releasing a deadlock.

@@

To each safe step $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$ coming with a network configuration $\sigma$ and a direction map $d$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$, next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\sigma$ during the initial step  and we lock signals to red for each train's destination according to $d$. 
\begin{itemize}
\item for each train $\tid{i}$, $\trainSeq[\tid{i}] = \trainTuple{\tid{i}}{x_1^{\tid{i}}}{\dirStop}{\su{d(\tid{i})}{\vec{x}^{\tid{i}}}}$,
\item $R = \langle E,  \bid{i} \in \blocks \mapsto 0, W\rangle$, $E$ being defined just below,
\item $W = \{ \tuple{x_1^\tid{i}, x_1^\tid{i}, 1, \tid{i}}  \mid \tid{i} \in \trains \}$, 
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$: 
$$\langle \bid{i}, d\rangle \mapsto \left \{ \begin{array}{ll}
\sigred & \text{ if } \bid{i} = x_1^\tid{j}  \text{ for some }j \in \trains\\%  \\
\siggreen & \text{otherwise}
\end{array}\right.
$$
\end{itemize}

Now, let us turn to the definition of $E$. For any train $\tid{i} \in \trains$, we define $A= (\emptyList)_{x \in \vec{x}^\tid{i} }$. Then, set  $E[\tid{i}] = \langle d(\tid{i}), A[x_1 \leftarrow \langle \incr{x_1^\tid{i}}] \rangle]$.

@@

Two states $S$ and $S'$ are said to be \emph{compatible} whenever  for any train $\tid{i}$, we have $\mathtt{dst}(S, \tid{i})  =  \mathtt{src}(S', \tid{i})$. In other words, trains must start in $S'$ at destinations of $S$. We define the composition only compatible states. Actually, since we are interested by safe steps compositions, we restrict our attention to their compositions.

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}$. Let $S = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \regulator, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \regulator', \switches, \signals\rangle \rangle$ as follows. First,  for any train $\tid{i} \in \trains$, let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$.

Second, to define $E'$, we start $E' = E$ and we enrich it as follows.  For each
